---
title: "04_TFActivity_cleaned_up"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Info

Fourth step: estimate transcription factor (TF) activity from transcriptomics
data using **DoRothEA**.

## Getting Started

Load the required libraries and the support functions.

```{r libraries, message=FALSE}
library(progeny)
library(dorothea)
library(tibble)
library(tidyr)
library(dplyr)
library(ggplot2)
library(pheatmap)
library(readr)

## For the volcano plot (related to support functions)
library(ggrepel)

## We also load the support functions
source("support_functions.R")

```

## TF Activity with Dorothea

First, we read the original data and the result from the previous steps: 
+ Normalised transcriptomics data from (original data)
+ Differential expression analysis results (step 2). 
In this particular case, a top table results object from limma.

Then the format is modified, and the TF activity can be estimated with the help
of Dorothea. 

```{r loadInput, message=FALSE}

setwd("C:/Users/20192234/Documents/TUe/BMT Y3/8QC00-Tcells/data_projectTcell")

memory = FALSE
if (memory == TRUE){
  #memory
  x = c("Th0vsResting_memory.csv",
      "Th1vsResting_memory.csv",
      "Th2vsResting_memory.csv",
      "Th17vsResting_memory.csv",
      "iTregvsResting_memory.csv",
      "IFNBvsResting_memory.csv")
  Normalised_counts <- read_csv("../data_projectTcell/count_memory_df_vsn.csv")
  Experimental_design <- read_csv("../support_project/targets_memory.csv")
} else if (memory == FALSE){
  #naive
  x = c("Th0vsResting_naive.csv",
      "Th1vsResting_naive.csv",
      "Th2vsResting_naive.csv",
      "Th17vsResting_naive.csv",
      "iTregvsResting_naive.csv",
      "IFNBvsResting_naive.csv")
  Normalised_counts <- read_csv("../data_projectTcell/count_naive_df_vsn.csv")
  Experimental_design <- read_csv("../support_project/targets_naive.csv")
}

TFs_top_all_cond <- list()
counter = 1

for (i in x) {
  
  ## We read the results from the differential analysis. 
  ttop_filename = paste("ttop", i, sep="_")
  ttop_filepath = "../results_project/"
  ttop_ivsResting <- read_csv(paste(ttop_filepath, ttop_filename, sep=""))
 
  # We have to slightly modify the format of the input files to make them suitable
  # for running **DoRothEA**. 

  Normalised_counts_matrix <- Normalised_counts %>% 
      dplyr::mutate_if(~ any(is.na(.x)),~ if_else(is.na(.x),0,.x)) %>% 
      tibble::column_to_rownames(var = "gene") %>% 
      as.matrix()
  
  ttop_ivsResting_matrix <- ttop_ivsResting %>% 
      dplyr::select(ID, t) %>% 
      dplyr::filter(!is.na(t)) %>% 
      column_to_rownames(var = "ID") %>%
      as.matrix()

  
  ##now we start with the actual Dorothea part:
  
  # We estimate the transcription factor (TF) activity using the DoRothEA R 
  # package. We select interactions with confidence level A, B and C. 
  
  
  ## We load Dorothea Regulons
  data(dorothea_hs, package = "dorothea")
  regulons <- dorothea_hs %>%
    dplyr::filter(confidence %in% c("A", "B","C"))


  # Note that for DoRothEA, we proceed the other way around than for 
  # PROGENy. We have many TFs, so we cannot clearly visualize all of them in the 
  # same heatmap. That is why we first compute a TF activity enrichment analysis 
  # using the statistics from the differential expression analysis. This will allow 
  # us to select the TFs whose activity varies with the conditions under study. 
  # 
  # It is important to set the parameter `eset.filter` to `FALSE`. In this case,
  # we set the minimum size of regulons to five (`minsize`). I strongly recommend
  # to check `?viper` to set the parameters that best fit your needs. 
  

  tf_activities_stat <- dorothea::run_viper(ttop_ivsResting_matrix, regulons,
      options =  list(minsize = 5, eset.filter = FALSE, 
      cores = 1, verbose = FALSE, nes = TRUE))

  #for later purposes - all TFs are saved for each condition, later the average top 30 will be determined 
  tf_activities_stat_topall <- tf_activities_stat %>%
      as.data.frame() %>% 
      rownames_to_column(var = "GeneID") %>%
      dplyr::rename(NES = "t") %>%
      #dplyr::top_n(50, wt = abs(NES)) %>%
      dplyr::arrange(NES) %>% 
      dplyr::mutate(GeneID = factor(GeneID)) 
  TFs_top_all_cond[[counter]] = tf_activities_stat_topall
  counter = counter+1
  
  #tf_activities_stat_topall is matrix: 
  #first column: GeneID with names of all TFs
  #second column: NES of all TFs
  #important note: NES should be analysed for absolute numbers
  
  # The TF activity enrichment results provided by **Viper** are used as an input
  # in the **CARNIVAL** method. **CARNIVAL** tries to infer the most likely 
  # upstream signaling events leading to the current TF activity results. 
  
  #tf_activities_CARNIVALinput<- tf_activities_stat %>%
      #as.data.frame() %>% 
      #tibble::rownames_to_column(var = "TF") 

  #result_filename = paste("TFActivity_CARNIVALinput", i, sep = "_")
  #result_filepath = "../results_project_test_forloop/"

  #write_csv(tf_activities_CARNIVALinput, paste(result_filepath, result_filename, sep=""))
} 
```

## Determining the top TFs 

The top 30 TFs will be found in the following way: 
A matrix is set up with all TF Activity for all conditions, each row will correspond 
to a specific TF. The TFs will be ranked for each condition, these ranks will be 
averaged for all conditions for each TF. From these averaged ranks, the 30 most 
significant TFs can be found.

```{r determine TFs}
#create matrix as described by Óscar
#columns: diff conditions
#rows: TFs
#values in matrix: ranks for TF for specific condition

conditions = c("Th0",
      "Th1",
      "Th2",
      "Th17",
      "iTreg",
      "IFNB")
rank_matrix = as.data.frame(matrix())

#first: create rank column per condition; keep absolute values in mind!
for (i in 1:length(conditions)) {
  topTFs <- TFs_top_all_cond[[i]] %>%
      as.data.frame() %>% 
      #rownames_to_column(var = "GeneID") %>%
      #dplyr::rename(NES = "t") %>%
      #dplyr::top_n(35, wt = abs(NES)) %>%
      dplyr::arrange(abs(NES)) #%>% #desc() should give it in descending order, but this gives error?
      #dplyr::mutate(GeneID = factor(GeneID)) 
  
  rank_column = as.data.frame(matrix(nrow = length(topTFs[,1]), ncol = 2))
  rank_column[,1] = topTFs[,1]
  colnames(rank_column) = c("GeneID",conditions[i])
  for (j in 1:length(topTFs[,1])) {
    ranknr = (length(topTFs[,1])+1)-j
    rank_column[j,2] = ranknr
    
  }
  
  if (i == 1) {
    rank_matrix = rank_column  
  }
  else {
    rank_matrix = merge(rank_matrix, rank_column, by='GeneID')  
  }
}

#second: find average rank for each TF with rowMeans()
averageranks = rowMeans(rank_matrix[,2:ncol(rank_matrix)])
rank_matrix[,ncol(rank_matrix)+1] = averageranks
colnames(rank_matrix)[ncol(rank_matrix)] = "average"

#third: select top n TFs 
#1: reorder rank_matrix
rank_matrix_reordered = arrange(rank_matrix, rank_matrix[,8])
top30TFs_conditions_combined = head(rank_matrix_reordered, n = 30) 

#only thing to do now: 
#write new .csv file with NES for all TFs for all conditions (see next block)

```

## Saving data in a .csv file format 

The rows of the data in the .csv file correspond to the TFs, the columns to 
the conditions. 

```{r write files with info}
#where to find NESs of all conditions: TFs_top_all_cond[[i]]
conditions = c("Th0",
      "Th1",
      "Th2",
      "Th17",
      "iTreg",
      "IFNB")

tf_activities_top30 = top30TFs_conditions_combined[, 1] %>% as.data.frame()
names(tf_activities_top30)[1] <- "GeneID"
for (i in 1:length(conditions)) {
  tf_activities_top30 = merge(tf_activities_top30, TFs_top_all_cond[[i]], by='GeneID')
  names(tf_activities_top30)[i+1] <- conditions[i]
}

if (memory == TRUE) {
  #memory
  result_filename = paste("TFActivity_CARNIVALinput", "memory", sep = "_")
} else {
  #naive
  result_filename = paste("TFActivity_CARNIVALinput", "naive", sep = "_")
}
result_filepath = "../00cleaned_up_files/results_cleaned_up/"

write_csv(tf_activities_top30, paste(result_filepath, result_filename, sep=""))

```


