---
title: "05_running_CARNIVAL_cleaned_up"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro

Fifth step: running CARNIVAL with the data from steps 3 and 4. 

## Getting Started

Load the required libraries and the support functions.

```{r libraries, message=FALSE}
library(progeny)
library(dorothea)
library(CARNIVAL)
library(OmnipathR)
library(readr)
library(tibble)
library(tidyr)
library(dplyr)
library(visNetwork)

library(ggplot2)
library(pheatmap)


## For the volcano plot (related to support functions)
library(ggrepel)

## We also load the support functions
source("assignPROGENyScores.r")
source("generateTFList.r")
source("carnival_visNetwork.r")
```

In addition, read the results from the previous steps: 

+ Transcription factor activities (step 4)
+ Pathways activity scores (step 3). 

```{r loadInput, message=FALSE}
setwd("C:/Users/20192234/Documents/TUe/BMT Y3/8QC00-Tcells/scripts_project/00cleaned_up_files/results_cleaned_up")

## We read the normalised counts and the experimental design 

memory = FALSE

if (memory == FALSE) {
  ## naive
  tf_activities <- read_csv("../results_cleaned_up/TFActivity_CARNIVALinput_naive")
  PathwayActivity <- read_csv("../results_cleaned_up/PathwayActivity_CARNIVALinput_naive")
} else {
  ## memory
  tf_activities <- read_csv("../results_cleaned_up/TFActivity_CARNIVALinput_memory")
  PathwayActivity <- read_csv("../results_cleaned_up/PathwayActivity_CARNIVALinput_memory")
}

```

## Getting the scaffold network from original data

Instead of using OmniPath to get the scaffold network (aka the pkn), we load in 
a curated network based on papers about T cell networks. 

```{r omnipathSIF, message=FALSE}
setwd("C:/Users/20192234/Documents/TUe/BMT Y3/8QC00-Tcells/data_projectTcell")

sif <- read.table("../data_projectTcell/curated_network_CARNIVALinput.sif")
colnames(sif) <- c("source", "interaction", "target")

#save SIF
write_tsv(sif, "../data_projectTcell/curated_network_CARNIVALinput.tsv")

```

## Transcription Factor and pathway activities for CARNIVAL

We use the supplementary function _generateTFList.r_ to shift the format of 
tf_activities to the one required by CARNIVAL. The format of 
PathwayActivity_carnival is also adapted. 

```{r shiftFormats, message=FALSE}

# dorothea for CARNIVAL
tf_activities_carnival <- data.frame(tf_activities, stringsAsFactors = F, row.names = 1)
ttf_activities_carnival <- t(tf_activities_carnival)
#rows: conditions; columns: tfs

#30 TFs are selected:
tfList = generateTFList(tf_activities_carnival, top=30, access_idx = 1:6)
#warning: tfList now still contains data for ALL conditions

# progeny for CARNIVAL
PathwayActivity_carnival <- data.frame(PathwayActivity, stringsAsFactors = F, row.names = 1)
tPathwayActivity_carnival <- as.data.frame(t(PathwayActivity_carnival))
#rows: conditions; columns: pathways

conditions = c("Th0",
      "Th1",
      "Th2",
      "Th17",
      "iTreg",
      "IFNB")

pxList = split(tPathwayActivity_carnival, conditions)

```

## Running CARNIVAL

CARNIVAL has been developed to find the causal link between the activities of the transcription factors (TFs) and the 'perturbed' nodes.
We have 3 main inputs that we have to provide:

+ _measObj_: The TFs' activities (like the ones we have obtained from DoRothEA)
+ _inputObj_: The 'perturbed' nodes we want that CARNIVAL connects with the activity of TFs. 
There are 3 ways of using it:
(1) Give the name and sign of the selected nodes; 
(2) Give the name only, so the algorithm will select the sign that best fit the models,
(3) Give _NULL_ as value will create a "Perturbation" node that will try both signs for all 'initial' nodes of the given network ( _netObj_ ).
+ _netObj_: The network that will serve as map to connect the TFs' activities ( _measObj_ ) and the perturbed nodes ( _inputObj_ )

Although it is not required, a fourth object called _weightObj_ can be also given. 
This object gives values ranged from -1 to 1 for a set of nodes of the network.
The aim of _weightObj_ is helping the solver to find optimal solutions faster.

```{r prep, message= FALSE}
#prep for iniciator matrix
# MC
iniMTX_distinct_ch = base::data.frame("IL12_EX","IL4_EX", "TGFB_EX", "IL6_EX", "IL23_EX", "IL1B_EX", "IL2_EX", "IFNB_EX", "CD28", "TCR")
iniciators = base::data.frame(base::matrix(data = 0, nrow = 6, ncol = length(iniMTX_distinct_ch)), stringsAsFactors = F)
colnames(iniciators) = iniMTX_distinct_ch

#Th0 (but also for all the cell states)
iniciators["CD28"] <- 1
iniciators["TCR"] <- 1

#Th1
iniciators[2,"IL12_EX"] <- 1

#Th2
iniciators[3, "IL4_EX"] <- 1

#Th17
iniciators[4, "TGFB_EX"] <- 1
iniciators[4, "IL6_EX"] <- 1
iniciators[4, "IL23_EX"] <- 1
iniciators[4, "IL1B_EX"] <- 1

#Treg
iniciators[5, "TGFB_EX"] <- 1
iniciators[5, "IL2_EX"] <- 1

#IFNB
iniciators[6, "IFNB_EX"] <- 1

```

```{r carnival, message=FALSE}
#we want to run carnival for all conditions, so we create a forloop:
#important:
#iniciators --> perturbed nodes (starting points)
#tPathwayActivity_carnival --> pathway info (weights)
#ttf_activities_carnival --> tfs info (ending points)
#sif --> pkn

if (memory == FALSE) {
  #naive
  result_filename = paste("CARNIVAL_result", "naive", sep = "_")
}else {
  #memory
  result_filename = paste("CARNIVAL_result", "memory", sep = "_")
}
result_filepath = "../00cleaned_up_files/results_cleaned_up//"

conditions = c("Th0",
      "Th1",
      "Th2",
      "Th17",
      "iTreg",
      "IFNB")

for (i in 1:length(conditions)){
  #get right info
  cond_ini <- iniciators %>% slice(i:i)
  
  if (conditions[i] == "Th0") {
    cond_TFActivity <- tfList$Th0
    cond_PathwayActivity <- pxList$Th0
  } else if (conditions[i] == "Th1") {
    cond_TFActivity <- tfList$Th1
    cond_PathwayActivity <- pxList$Th1
  } else if (conditions[i] == "Th2") {
    cond_TFActivity <- tfList$Th2
    cond_PathwayActivity <- pxList$Th2
  } else if (conditions[i] == "Th17") {
    cond_TFActivity <- tfList$Th17
    cond_PathwayActivity <- pxList$Th17
  } else if (conditions[i] == "iTreg") {
    cond_TFActivity <- tfList$iTreg
    cond_PathwayActivity <- pxList$iTreg
  } else if (conditions[i] == "IFNB") {
    cond_TFActivity <- tfList$IFNB
    cond_PathwayActivity <- pxList$IFNB
  }
  
  save_genes <- names(cond_PathwayActivity)
  cond_PathwayActivity <- as.numeric(cond_PathwayActivity)
  names(cond_PathwayActivity) <- save_genes
  
  #run carnival
  carnival_result = runCARNIVAL(#changes for every condition (except netObj)
                                inputObj= cond_ini,
                                measObj = cond_TFActivity,
                                netObj = sif,
                                weightObj = cond_PathwayActivity,
                               
                                #keep this the same
                                solverPath = "C:/Program Files/IBM/ILOG/CPLEX_Studio201/cplex/bin/x64_win64",
                                solver = "cplex",
                                timelimit=120,
                                mipGAP=0,
                                poolrelGAP=0,
                                )
  #save solution 
  carnival_result$weightedSIF <- data.frame(carnival_result$weightedSIF, stringsAsFactors = F)
  carnival_result$weightedSIF$Sign <- as.numeric(carnival_result$weightedSIF$Sign)
  carnival_result$weightedSIF$Weight <- as.numeric(carnival_result$weightedSIF$Weight)

  carnival_result$nodesAttributes <- data.frame(carnival_result$nodesAttributes, stringsAsFactors = F)
  carnival_result$nodesAttributes$ZeroAct <- as.numeric(carnival_result$nodesAttributes$ZeroAct)
  carnival_result$nodesAttributes$UpAct <- as.numeric(carnival_result$nodesAttributes$UpAct)
  carnival_result$nodesAttributes$DownAct <- as.numeric(carnival_result$nodesAttributes$DownAct)
  carnival_result$nodesAttributes$AvgAct <- as.numeric(carnival_result$nodesAttributes$AvgAct)
  
  result_filename_final = paste(result_filename, conditions[i], sep = "_")
  saveRDS(carnival_result, paste(result_filepath, result_filename_final, sep=""))
  
}


```



