---
title: "06_analysis_CARNIVAL_results"
author: "Rosa Hernansaiz-Ballesteros"
date: "14/08/2020"
output: 
  github_document: default
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### License Info

This program is free software: you can redistribute it and/or modify it under 
the terms of the GNU General Public License as published by the Free Software 
Foundation, either version 3 of the License, or (at your option) any later 
version.

This program is distributed in the hope that it will be useful, but WITHOUT ANY 
WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR 
A PARTICULAR PURPOSE. See the GNU General Public License for more details.

Please check http://www.gnu.org/licenses/.

## Introduction
This is the final part of our series of transcriptomics tutorials. In previous parts, we normalised RNASeq data for differential analysis, and used the differential analysis results in **PROGENy** and **DOROTHea**. In the previous tutorial, we demonstrated **CARNIVAL** with TF activities from **DOROTHea**, a prior knowledge network from **Omnipath**, and weights from **PROGENy**.

## CARNIVAL output

CARNIVAL (CAusal Reasoning for Network identification using Integer VALue programming) 
returns a list with 4 elements: 

* weightedSIF. Character matrix. Contains a summary of _sifAll_, 
where *Weight* indicates the number of networks that present that interaction.

* nodesAttributes. Character matrix. Contains a summary of _attributesAll_, 
where all columns ending in *Act* indicate the number of networks that present that node in
each of the states (Zero = doesn't appear in the network, Up = upregulated, Down = downregulated, Avg = average regulation).
The column *NodeType* indicates if the node is a TF's (T) or a perturbed one (S).

* sifAll. List of character matrices. All networks that have been calculated.

* attributesAll. List of character matrices.

NOTE: This is valid for current CARNIVAL version (1.0.0)

## CARNIVAL interpretation

The summary files ( _sifAll_ and _attributesAll_ ) can be directly used to identify causal interactions between the 
perturbed nodes and the selected Transcription Factors. 
In addition to extracting direct information from the network, 
we can run different downstream analysis based on the necessities of each project.

Here are described some of the downstream approaches that we have used:

+ Pathway enrichment analysis
+ Network comparisons

# Get ready

```{r libraries, message=FALSE}
library(readr)
library(piano)
library(dplyr)
library(ggplot2)
library(tibble)
library(tidyr)
library(dplyr)
library(scales)
library(plyr)
library(GSEABase)
library(network)
library(reshape2)
library(cowplot)
library(pheatmap)
library(ggraph)
library(tidygraph)

# set working directory
setwd("C:/Users/20182400/Documents/Rstudio OGO git/8QC00-Tcells/scripts_project")

## We also load the support functions
source("support_enrichment.r")
source("support_networks.r")

## Read the data
pkn = read_tsv("../results_project/network_oscar.tsv")

```

# Topological measurments

We can get a sense of the size of the reconstructed networks using topological parameters, such as number of edges and nodes.
We can also have a look at the network's density and the *degree distribution*.
The *density* indicates the proportion of interactions that exist in our network when comparing with all possible interactions that can be established. 
The *degree distribution* shows the number of connections of a node. 
In a directed network, we can distinguish between incoming and outgoing connections.

NOTE: Here the density is calculated for a _directed graph_.
As CARNIVAL can report 2 interactions between the same 2 nodes with different sign, 
these "doubled" interactions are excluded when calculating the density.

For this section, and the following ones, we use the CARNIVAL networks obtained
when running the _sample_resolution_carnival.R_. 

To find out more about this topic:

+ https://mathinsight.org/degree_distribution
+ https://www.networksciencebook.com/chapter/2#degree

```{r netopology, message=FALSE, results='hide', warning=FALSE}

#Get results memory
carnival_result_Th0_memory <- readRDS("~/Rstudio OGO git/8QC00-Tcells/scripts_project/00cleaned_up_files/results_cleaned_up/CARNIVAL_result_memory_Th0.rds")
carnival_result_Th1_memory <- readRDS("~/Rstudio OGO git/8QC00-Tcells/scripts_project/00cleaned_up_files/results_cleaned_up/CARNIVAL_result_memory_Th1.rds")
carnival_result_Th2_memory <- readRDS("~/Rstudio OGO git/8QC00-Tcells/scripts_project/00cleaned_up_files/results_cleaned_up/CARNIVAL_result_memory_Th2.rds")
carnival_result_Th17_memory <- readRDS("~/Rstudio OGO git/8QC00-Tcells/scripts_project/00cleaned_up_files/results_cleaned_up/CARNIVAL_result_memory_Th17.rds")
carnival_result_IFNB_memory <- readRDS("~/Rstudio OGO git/8QC00-Tcells/scripts_project/00cleaned_up_files/results_cleaned_up/CARNIVAL_result_memory_IFNB.rds")
carnival_result_iTreg_memory <- readRDS("~/Rstudio OGO git/8QC00-Tcells/scripts_project/00cleaned_up_files/results_cleaned_up/CARNIVAL_result_memory_iTreg.rds")
#Add together memory
carnival_results_list_memory <- list(Th0_memory = carnival_result_Th0_memory,Th1_memory = carnival_result_Th1_memory, Th2_memory = carnival_result_Th2_memory, Th17_memory = carnival_result_Th17_memory, IFNB_memory = carnival_result_IFNB_memory, iTreg_memory = carnival_result_iTreg_memory)

#Get results naive
carnival_result_Th0_naive <- readRDS("~/Rstudio OGO git/8QC00-Tcells/scripts_project/00cleaned_up_files/results_cleaned_up/CARNIVAL_result_naive_Th0.rds")
carnival_result_Th1_naive <- readRDS("~/Rstudio OGO git/8QC00-Tcells/scripts_project/00cleaned_up_files/results_cleaned_up/CARNIVAL_result_naive_Th1.rds")
carnival_result_Th2_naive <- readRDS("~/Rstudio OGO git/8QC00-Tcells/scripts_project/00cleaned_up_files/results_cleaned_up/CARNIVAL_result_naive_Th2.rds")
carnival_result_Th17_naive <- readRDS("~/Rstudio OGO git/8QC00-Tcells/scripts_project/00cleaned_up_files/results_cleaned_up/CARNIVAL_result_naive_Th17.rds")
carnival_result_IFNB_naive <- readRDS("~/Rstudio OGO git/8QC00-Tcells/scripts_project/00cleaned_up_files/results_cleaned_up/CARNIVAL_result_naive_IFNB.rds")
carnival_result_iTreg_naive <- readRDS("~/Rstudio OGO git/8QC00-Tcells/scripts_project/00cleaned_up_files/results_cleaned_up/CARNIVAL_result_naive_iTreg.rds")
#Add together naive
carnival_results_list_naive <- list(Th0_naive = carnival_result_Th0_naive,Th1_naive = carnival_result_Th1_naive, Th2_naive = carnival_result_Th2_naive, Th17_naive = carnival_result_Th17_naive, IFNB_naive = carnival_result_IFNB_naive, iTreg_naive = carnival_result_iTreg_naive)

# get only summary files from CARNIVAL results
sifts = lapply(carnival_results_list_memory, function(x){x$weightedSIF}) 
nodos = lapply(carnival_results_list_memory, function(x){x$nodesAttributes})

sifts1 = lapply(carnival_results_list_naive, function(x){x$weightedSIF}) 
nodos1 = lapply(carnival_results_list_naive, function(x){x$nodesAttributes})

# Calculate the number of edges and nodes in the networks and its density
node_edge = do.call(rbind,lapply(sifts, count_edges_nodes_degree))
node_edge1 = do.call(rbind,lapply(sifts1, count_edges_nodes_degree))

# Calculate degree distribution for a sample
count_degree = sifts$IFNB_memory %>% degree_count 
count_degree1 = sifts1$IFNB_naive %>% degree_count

#visualise
ggdat = as.data.frame(node_edge) %>% tibble::rownames_to_column(var = "sample") %>%
  dplyr::mutate(condition = gsub(".Rep[0-9]{1}", "", sample))
ggdat1 = as.data.frame(node_edge1) %>% tibble::rownames_to_column(var = "sample") %>%
  dplyr::mutate(condition = gsub(".Rep[0-9]{1}", "", sample))

ggdat_total <- rbind(ggdat, ggdat1)

#Plotting
nodes <- ggdat_total[,"nodes"]
edges <- ggdat_total[,"edges"]

# relation between number of edges and nodes
ggplot(ggdat_total, aes(x = nodes, y = edges, color = as.factor(condition))) +
  geom_point() +
  xlim(0 , 25) +
  ylim(0 , 25) +
  geom_text(label = ggdat_total$sample, 
            check_overlap = TRUE,
            vjust = 0, 
            nudge_y = 0.5,
            show.legend = F,
            mapping = NULL,
            size = 3.5) +
  theme_bw(base_size = 14) +
  guides(color = guide_legend( title="Conditions" ) ) +
  ggtitle("Node-edge composition")

ggsave("00cleaned_up_files/results_cleaned_up/node_edge_composition.pdf")


#network degree
ggplot(ggdat_total, aes(x = density, y = sample,fill = as.factor(condition))) + 
  geom_point() +
  geom_text(
    aes(label = sprintf("%0.3f", density)), 
    position = position_dodge(width = 1),
    inherit.aes = TRUE,
    hjust = -0.1, size = 4,
    ) +
  xlim(0 , 0.15) +
  geom_col() +
  theme_bw(base_size = 14) +
  guides(fill = "none" ) + 
  ggtitle("Network degree")

ggsave("00cleaned_up_files/results_cleaned_up/network_degree.pdf")
```


Generally speaking, biological networks are not dense, so don't worry if the density values are low... they usually are!

The degree distribution is an interesting graphic to look at, as we can get some immediate information:

+ k = 0 for in-degree indicates the proportion of initial nodes, 
while for out-degree indicates the effectors (here the TFs). 
For total-degree, that would indicate the presence of isolated nodes.

+ Most of the nodes do not have many connections (e.g. k = 1 -> p = 0.6), 
but there are a few that are highly connected (e.g. k > 6).

+ We can find some hubs when k is higher (plot B). The out-degree ends at k = 5; 
this means that the biggest regulatory-hub regulates at most 5 other nodes.
In a similar way, the in-degree goes up to k = 9;
This means that there are few hubs (k > 5) that are regulated by at most 9 nodes.

# Network comparison

When we have more than one network, we usually would like to know how (dis)similar these networks are.

We can use the *Jaccard Index* to measure similarities and diversity
between sample sets (https://en.wikipedia.org/wiki/Jaccard_index).Int 
This index can be used in different scenarios.
Generally, it's used with the nodes, but we can also use it with the edges.

When we have groups, in our case WT and FOXA2KO, we can also extract the interactions that are common.

```{r netcompare, message=FALSE, results='hide', warning=FALSE}

# create a matrix of all interactions for all samples
interactions = getTopology(networks = sifts, scafoldNET = pkn)
interactions1 = getTopology(networks = sifts1, scafoldNET = pkn)

# get the edges per sample
net_int = apply(interactions, 2, function(x, r){
  r[which(!is.na(x))]
}, rownames(interactions))
net_int1 = apply(interactions1, 2, function(x, r){
  r[which(!is.na(x))]
}, rownames(interactions1))

# calculate Jaccard indexes per pair
combined = expand.grid(1:length(names(sifts)), 1:length(names(sifts1)))
jac_index = matrix(data = NA, nrow = length(names(sifts)), ncol = length(names(sifts1)),
                   dimnames = list(names(sifts), names(sifts1)))

for (i in 1:nrow(combined)){
  n = names(sifts)[combined[i,1]]
  m = names(sifts1)[combined[i,2]]
  jac_index[n,m] = length( intersect(net_int[[n]], net_int1[[m]]) ) / length( union(net_int[[n]], net_int1[[m]]) )
}

#Visualize the indices in a heatmap

#saving method
save_pheatmap_pdf <- function(x, filename, width=7, height=7) {
   stopifnot(!missing(x))
   stopifnot(!missing(filename))
   pdf(filename, width=width, height=height)
   grid::grid.newpage()
   grid::grid.draw(x$gtable)
   dev.off()
}

heatmap_ji <- pheatmap(jac_index,fontsize=14, 
           fontsize_row = 11, fontsize_col = 11, 
           angle_col = 45, treeheight_col = 0, cellheight = 40, cellwidth = 50,
           main = "Jaccard indices")

save_pheatmap_pdf(heatmap_ji, "C:/Users/20182400/Documents/RStudio OGO git/8QC00-Tcells/scripts_project/00cleaned_up_files/results_cleaned_up/jaccard_index.pdf")

```

## Session Info Details

```{r sessionInfo, echo=FALSE, eval=TRUE}
sessionInfo()
```
